---
title: "01_make_data_PP"
author: "Praveen Puviindran"
date: "2025-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Describe the project
This project is analyzing a synthetic lung cancer dataset containing numerous variables including patient demographics, tumor characteristics, treatment approaches, and biological data to hopefully be able to identify patterns or key factors that influence lung cancer prognosis and treatment efficacy.

# Killer graph description
What I believe would be considered a standalone "killer graph" for this project would be some sort of feature importance graph that incorporates all of the various measured variables and highlights the most influential ones in predicting cancer prognosis and treatment efficacy. This could be in combination with an overlayed density map or bar graph that shows the proportion of influence a variable has on the outcome.

```{r, echo=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(DT)
library(GGally)
library(tibble)
library(gridExtra)
```

```{r}
# Load the data
file_path <- "/Users/praveenpuviindran/Documents/Current Course Material/DATA481S25/lung_cancer_data.csv"
lung_cancer_data <- read.csv(file_path)

# Preview the data
head(lung_cancer_data)

# Summary of the dataset
summary(lung_cancer_data)

# Check for missing values
tibble(Missing_Count = colSums(is.na(lung_cancer_data)))
```

### Continuous Variable Distributions
```{r}
# Plot distributions of continuous variables
continuous_vars <- c("Age", "Tumor_Size_mm", "Survival_Months")
plots <- list()
for (var in continuous_vars) {
  plots[[var]] <- ggplot(lung_cancer_data, aes_string(x = var)) +
    geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste("Distribution of", var), x = var, y = "Count")
}
grid.arrange(grobs = plots, ncol = 2)
```

### Relationships Between Variables
```{r}
# Scatterplots for numerical relationships
scatter_plot <- ggplot(lung_cancer_data, aes(x = Tumor_Size_mm, y = Survival_Months)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Tumor Size vs Survival Months", x = "Tumor Size (mm)", y = "Survival Months")

scatter_plot
```

### Boxplots for Survival by Categorical Variables
```{r}
# Boxplots for Survival_Months by categorical variables
boxplot_vars <- c("Stage", "Treatment", "Tumor_Location")
plots <- list()
for (var in boxplot_vars) {
  plots[[var]] <- ggplot(lung_cancer_data, aes_string(x = var, y = "Survival_Months", fill = var)) +
    geom_boxplot() +
    theme_minimal() +
    labs(title = paste("Survival Months by", var), x = var, y = "Survival Months") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
grid.arrange(grobs = plots, ncol = 2)
```

### Pair Plot for Numerical Variables
```{r}
# Pair plot for numerical variables
continuous_subset <- lung_cancer_data %>% select(Age, Tumor_Size_mm, Survival_Months)
ggpairs(continuous_subset)
```

### Correlation Analysis
```{r}
# Filter only numeric columns and exclude irrelevant or ID columns
correlation_data <- lung_cancer_data %>%
  select_if(is.numeric)

# Calculate correlation matrix
correlation_matrix <- round(cor(correlation_data, use = "complete.obs"), 2)

# Melt the correlation matrix for visualization
melted_correlation <- melt(correlation_matrix)

# Recreate the heatmap with clearer scales
heatmap_plot <- ggplot(melted_correlation, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

heatmap_plot
```

# Killer Graph
The "killer graph" for this project would be a hybrid visualization combining a feature importance chart and Kaplan-Meier survival curves. The feature importance chart would rank the key factors, such as cancer stage, tumor size, treatment type, and smoking history, based on their influence on lung cancer prognosis and survival outcomes, as determined by a machine learning model. The Kaplan-Meier survival curves would provide deeper insights by showing how survival probabilities change over time for specific groups, such as patients at different cancer stages or undergoing different treatments. By integrating these two elements, the graph would highlight both the statistical importance of each factor and how these factors translate into actual survival trends, offering a comprehensive view of what drives patient outcomes.

# Research Question
What are the key factors influencing survival months in lung cancer patients, and how do these factors vary across different stages and treatment types?

```{r}
# Filter data for each Tumor_Location and create separate plots
tumor_locations <- unique(lung_cancer_data$Tumor_Location)

plots <- list()

for (location in tumor_locations) {
  plot <- ggplot(lung_cancer_data %>% filter(Tumor_Location == location),
                 aes(x = Stage, y = Survival_Months, fill = Treatment)) +
    geom_boxplot() +
    theme_minimal() +
    labs(title = paste("Survival Months by Stage and Treatment -", location),
         x = "Stage", y = "Survival Months", fill = "Treatment")
  
  plots[[location]] <- plot
}

# Print each plot
for (location in tumor_locations) {
  print(plots[[location]])
}

```

```{r}
# Violin plot for Survival Months by Gender
ggplot(lung_cancer_data, aes(x = Gender, y = Survival_Months, fill = Gender)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_minimal() +
  labs(title = "Survival Months by Gender", x = "Gender", y = "Survival Months")

# Boxplot for Survival Months by Ethnicity
ggplot(lung_cancer_data, aes(x = Ethnicity, y = Survival_Months, fill = Ethnicity)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Survival Months by Ethnicity", x = "Ethnicity", y = "Survival Months")
```

```{r}
# Calculate median survival months by Stage and Treatment
treatment_summary <- lung_cancer_data %>%
  group_by(Stage, Treatment) %>%
  summarise(
    Median_Survival = median(Survival_Months, na.rm = TRUE),
    Mean_Survival = mean(Survival_Months, na.rm = TRUE),
    Count = n()
  )

# Print summary table
print(treatment_summary)

# Plot median survival by Treatment and Stage
ggplot(treatment_summary, aes(x = Treatment, y = Median_Survival, fill = Stage)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(
    title = "Median Survival Months by Treatment and Stage",
    x = "Treatment", y = "Median Survival Months", fill = "Stage"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Select all chemical/biological variables
bio_vars <- c("Hemoglobin_Level", "White_Blood_Cell_Count", "Platelet_Count", 
              "Albumin_Level", "Alkaline_Phosphatase_Level", "Alanine_Aminotransferase_Level", 
              "Aspartate_Aminotransferase_Level", "Creatinine_Level", "LDH_Level",
              "Calcium_Level", "Phosphorus_Level", "Glucose_Level", "Potassium_Level", 
              "Sodium_Level")

# Summary statistics
summary(lung_cancer_data[bio_vars])
```

```{r}
# Compute correlation matrix for biological markers
bio_corr_matrix <- cor(lung_cancer_data[bio_vars], use = "complete.obs")

# Convert to long format for heatmap
library(reshape2)
melted_corr <- melt(bio_corr_matrix)

# Heatmap visualization
ggplot(melted_corr, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Correlation Heatmap of Biological Markers", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r}
# Histogram for major biomarkers
p1 <- ggplot(lung_cancer_data, aes(x = Hemoglobin_Level)) +
  geom_histogram(fill = "blue", color = "black", bins = 30, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Hemoglobin Levels", x = "Hemoglobin Level", y = "Count")

p2 <- ggplot(lung_cancer_data, aes(x = White_Blood_Cell_Count)) +
  geom_histogram(fill = "red", color = "black", bins = 30, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of White Blood Cell Count", x = "WBC Count", y = "Count")

p3 <- ggplot(lung_cancer_data, aes(x = Albumin_Level)) +
  geom_histogram(fill = "green", color = "black", bins = 30, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Albumin Levels", x = "Albumin Level", y = "Count")

grid.arrange(p1, p2, p3, ncol = 2)
```

```{r}
# Boxplots for selected biomarkers across different tumor stages
p1 <- ggplot(lung_cancer_data, aes(x = Stage, y = Hemoglobin_Level, fill = Stage)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Hemoglobin Levels by Tumor Stage", x = "Stage", y = "Hemoglobin Level")

p2 <- ggplot(lung_cancer_data, aes(x = Stage, y = White_Blood_Cell_Count, fill = Stage)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "White Blood Cell Count by Tumor Stage", x = "Stage", y = "WBC Count")

p3 <- ggplot(lung_cancer_data, aes(x = Stage, y = Creatinine_Level, fill = Stage)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Creatinine Levels by Tumor Stage", x = "Stage", y = "Creatinine Level")

grid.arrange(p1, p2, p3, ncol = 2)
```

```{r}
# Scatterplots for biomarker vs survival months
p1 <- ggplot(lung_cancer_data, aes(x = Hemoglobin_Level, y = Survival_Months)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "Hemoglobin Level vs Survival Months", x = "Hemoglobin Level", y = "Survival Months")

p2 <- ggplot(lung_cancer_data, aes(x = White_Blood_Cell_Count, y = Survival_Months)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "WBC Count vs Survival Months", x = "WBC Count", y = "Survival Months")

p3 <- ggplot(lung_cancer_data, aes(x = Glucose_Level, y = Survival_Months)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "Glucose Level vs Survival Months", x = "Glucose Level", y = "Survival Months")

grid.arrange(p1, p2, p3, ncol = 2)
```

