---
title: "Lung Cancer Dataset Analysis"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tidyr)
library(gridExtra)
library(reshape2)

```

# Description of project
The project uses a synthetic data set on lung cancer patients for predictive modelling. The data set includes demographic attributes, medical history, treatment specifics, and survival outcomes, making it a great tool for simulating real-world clinical applications and to predict relationship between variables.

# Description of killer graph
To identify hidden and vital factors that has a statistically significant effect on survival outcomes in lung cancer patients. 

```{r}
lung_cancer_data <- read.csv("lung_cancer_data.csv")
head(lung_cancer_data)
```
```{r}
str(lung_cancer_data)
```
```{r}
summary(lung_cancer_data)
```
```{r}
cnrdata <- lung_cancer_data %>% select(-Patient_ID)
```

```{r}
# Examining distributions

ggplot(cnrdata, aes(x = Survival_Months)) +
  geom_histogram(binwidth = 5, fill = "skyblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Survival Months", x = "Survival Months", y = "Frequency") + theme_minimal()

ggplot(cnrdata, aes(x = Performance_Status)) +
  geom_histogram(binwidth = 5, fill = "skyblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Performance Status", x = "Performance Status", y = "Frequency") + theme_minimal()

ggplot(cnrdata, aes(x = Tumor_Size_mm)) +
  geom_histogram(binwidth = 5, fill = "skyblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Tumor Size", x = "Tumor Size", y = "Frequency") + theme_minimal()
```


```{r}
#Distribution of Categorical Variables

df_categorical <- cnrdata %>%
  select(Gender, Smoking_History, Stage, Treatment, Tumor_Location,  Comorbidity_Autoimmune_Disease)%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Category")

ggplot(df_categorical, aes(x = Category)) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free_x") +  
  labs(title = "Distribution of Categorical Variables", x = "Category", y = "Count") +  theme_minimal() + scale_x_discrete(guide = guide_axis(angle = 45))

```
```{r}
# #Distribution of Numerical Variables
continuous_vars <- c("Age",  "Smoking_Pack_Years", "Hemoglobin_Level", "Platelet_Count", "LDH_Level")

histograms <- lapply(continuous_vars, function(var) {
  ggplot(cnrdata, aes_string(x = var)) +
    geom_histogram(binwidth = 10, fill = "pink", color = "black", alpha = 0.7) +
    labs(title = paste("Distribution of", var), x = var, y = "Frequency") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10),
      plot.margin = margin(5, 5, 5, 5)  # Smaller plot margins
    )
})

# Arrange histograms in a grid layout with smaller spaces
do.call(grid.arrange, c(histograms, ncol = 2, top = "Histograms of Continuous Variables"))
```

```{r}
# Mean survival months vs. numerical variables
for (var in continuous_vars) {
  print(
    ggplot(cnrdata, aes_string(x = var, y = "Survival_Months")) +
      stat_summary_bin(
        fun = "mean", binwidth=5, geom = "bar", fill = "steelblue",
        alpha = 0.7 ) + labs( title = paste("Mean Survival Months by ", var), x = var, y = "Mean Survival Months") +theme_minimal(base_size = 14) +
      theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12)
      ))}

```

```{r}
# Heat map showing correlation between variables
numeric_data <- cnrdata %>% select(where(is.numeric))
corr_mat <- round(cor(numeric_data),2) 

melted_corr_mat = melt(corr_mat)

ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + 
geom_tile()+
theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
```

